---
title: "MultiSero Tutorial - Malaria Example"
author: "Madiha Shafquat"
date: "`r Sys.Date()`"
output:
 html_document:
  toc: true
  toc_float:
   toc_collapsed: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>"
)
```

Here we will use MultiSero to analyze data from the COMSA cross-sectional cohort in Mozambique.
Each individual in the dataset has antibody measurements against MSP1-19 antigens from _P. falciparum_, _P. malariae_, _P. ovale_, and _P. vivax_, measured on a multi-antigen Luminex assay.
Our primary outputs of interest are _Plasmodium_ species-specific seroprevalence and cross-reactivity estimates ($\phi$) for each combination of species.

# Set up
```{r message = FALSE, warning = FALSE}
# Required packages for this tutorial
library(here)           # File path management
library(data.table)     # Fast data manipulation
library(tidyverse)      # Data wrangling and visualization (includes ggplot2, dplyr, tidyr)
library(GGally)         # Extension to ggplot2 for pairwise plots
library(bayesplot)      # Visualization of Bayesian models and diagnostics

theme_set(theme_bw())

# source functions
source(here("R", "RFunctions.R"))
```

## Data pre-processing
For input into MultiSero, data should be in wide format, with one row per individual and one column per antigen.
The COMSA data has been pre-processed using BREAD and is available as normalized MFI values in long format.
t present, it is possible for BREAD to output MFIs that are below 0, but our model uses measurements on the log scale. Here we will only include inviduals with measurements > 0, but a different approach could also be appropriate depending on the dataset.
```{r}
# read in data
bread_raw <- read.csv(here("data", "BREAD_COMSA_output.csv")) |>
 filter(
  # only include antigens/pathogens of interest
  Antigen %in% c("pfmsp119_ls", "pmmsp119_ls", "pomsp119_ls", "pvmsp119_ls"),
  # keep strictly positive values
  Norm_MFI > 0
 )
head(bread_raw) |> knitr::kable()

df <- bread_raw |>
 # convert from long to wide format
 pivot_wider(
  id_cols = Sample, # sample ID
  names_from = Antigen, # antigen names become columns
  values_from = Norm_MFI # MFI values
 ) |>
 # rename antigen columns as desired to be more informative
 rename(
  P.falciparum = pfmsp119_ls,
  P.malariae = pmmsp119_ls,
  P.ovale = pomsp119_ls,
  P.vivax = pvmsp119_ls
 ) |>
 # remove rows with missing values created during pivoting/filtering
 drop_na()

head(df) |> knitr::kable()
```

Now that our data is ready, we define which pathogens are expected to be circulating in the population and are thus potential exposures.
In this setting, we expect _P. falciparium_, _P. malariae_, and _P. ovale_ to be present and _P. vivax_ to be absent.
If pathogen presence/absence is uncertain, we recommend a variable pathogen selection process (see [O'Driscoll et al.](https://doi.org/10.1126/scitranslmed.ads8680)).
```{r}
# define pathogen columns
pathogens <- colnames(df)[2:5]

present <- c("P.falciparum", "P.malariae", "P.ovale") # Pathogens assumed to be circulating
nonpres <- pathogens[!pathogens %in% present]         # Pathogens assumed absent

# reorder pathogen list with present pathogens first
pathogens <- c(present, nonpres)

# final data format
data <- format_data(
 df = df,
 present = present,
 nonpres = nonpres,
 ageG = FALSE,            # Age group indicator - TRUE if we want to stratify by age group
 locID = FALSE,           # Location indicator - TRUE if we want to stratify by location
 log_tranform = TRUE     # Whether titers need to be log-transformed
)
```

# Fit model
## Set priors
Before fitting, choose priors informed by the scale and distribution of your assay data, since multimodal posteriors are possible in this model.
```{r fig.width = 5}
ggpairs(log10(df[2:5]))
```

In this dataset, MFIs for all pathogens show a lower-density region near 3 and a higher peak close to 9 for _P. falciparum_, so we choose priors that are compatible with those ranges.
These priors are shared across pathogens. If diagnostics (see below section) suggest poor convergence, consider adjusting priors before refitting.

Priors for each model parameter follow a pre-specified distribution (see [O'Driscoll et al.](https://doi.org/10.1126/scitranslmed.ads8680)).
```{r}
priors <- list(
 mu0_mean = 3, mu0_sd = 0.1,    # Negative titers (normal distribution)
 mu1_mean = 3, mu1_sd = 0.5,    # Positive titers (normal distribution)
 sd0_mean = 1.5, sd0_sd = 0.05, # SD of negative titers (normal distribution)
 sd1_mean = 2, sd1_sd = 0.05,   # SD of positive titers (normal distribution)
 phi_rate = 8,                  # Cross-reactivity (exponential distribution)
 rho_a = 2, rho_b = 2           # Correlation (beta distribution)
)
```

## Set up cmdstanr
The model fitting is done in cmdstanr. Resources for setting up `cmdstanr` can be found [here](https://mc-stan.org/cmdstanr/articles/cmdstanr.html).
```{r}
library(cmdstanr)
# check cmdstan toolchain & set cmdstan path if needed
check_cmdstan_toolchain()
```

## Run MCMC
```{r}
# Compile the model
mod <- cmdstan_model(here("StanModels", "MultiSero_LocAge.stan"), pedantic = FALSE)

# Set output path
folder <- "comsa_mps119_res" # output subdirectory within Results/
```

```{r eval = FALSE}
dir.create(here("Results", folder))

# run model
fit <- mod$sample(
 data = c(data, priors),
 chains = 3,
 parallel_chains = 3,
 iter_sampling = 3000,
 refresh = 100,
 iter_warmup = 3000,
 output_dir = here("Results", folder)
)
```

```{r  include = FALSE}
# to avoid re-running, read in previously fit chains
output_dir <-  here("Results", folder)
fit <- as_cmdstan_fit(list.files(output_dir, pattern = "\\.csv$", full.names = TRUE))
```

## Check MCMC diagnostics
We use `bayesplot` diagnostics to check that chains mix well and converge to a common posterior distribution. Note that the estimated prevalence and $\mu_1$ for _P. vivax_ is set at 0 as it is not assumed to be present.
```{r fig.width = 9, fig.height = 5}
# extract posterior draws
chains <- fit$draws(format = "df")

trace_df <- relabel_chains(chains = chains, pathogens, present)

color_scheme_set("mix-blue-red")

# Examine trace plots for key parameter groups
mcmc_trace(trace_df, pars = vars(starts_with("Prev"), "Likelihood"))
mcmc_trace(trace_df, pars = vars(starts_with(c("Mu", "Std"))))
mcmc_trace(trace_df, pars = vars(starts_with("Phi"), "Rho"))
```
In this run, chains appear well mixed and converged. If not, increase iterations and/or revisit priors.

# Extract model estimates
Once diagnostics are acceptable, we can summarize posterior estimates.
```{r fig.width = 7, fig.height = 5}
posteriors <- get_posterior(as.data.frame(chains), data, pathogens)
posteriors$plot
posteriors$table |> knitr::kable(digits = 3)
```

Looking first at prevalence, estimated exposure is highest for _P. falciparum_ (median 0.840, 95% CrI 0.811-0.868), and much lower for the other two circulating strains: _P. malariae_ (0.158, 0.130-0.190) and _P. ovale_ (0.094, 0.073-0.119). 
These values can be interpreted as posterior probabilities of prior exposure in the sampled population, with credible intervals quantifying uncertainty.

The $\phi$ parameters describe how exposure to one species is expected to contribute to the antigen level of another speices.
We see modest cross-reactivity from exposure to _P. falciparum_ on the antibody level of _P. malariae_ ($\phi$ = 0.150 [0.131-0.168]), _P. ovale_ ($\phi$ = 0.131 [0.112-0.148]), and _P. vivax_ _P. ovale_ ($\phi$ = 0.131 [0.114-0.148]).